Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
more related work

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{| l |  l  | l | l | l | }
    \hline
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} \\
    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}

  \begin{tabular}{ll}
  \textbf{Features:} \\
  Universes: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
  Inductives: & Identity types (Id), Natural numbers (Nat), W-types~\cite{w-types} \\
  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
    \\
  \textbf{Citations:}\\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  NBE-in-Coq& \citet{nbeincoq} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
  Core Nuprl &\citet{anand2014towards} \\
  \end{tabular}

@@

In Figure~\ref{fig:featurematrix}, we give an overview of features of
the object languages from mechanized logical relations about dependent
types. The table is far from exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.

@@

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further shows it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion.
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.

@@

Finally, \citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous mechanization efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in Calculus of Constructions. We choose to
focus on a simple term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

There are other mechanized developments for dependently typed systems
that only involve properties that are derivable through syntactic
means. For example, \citet{coqcoqcorrect2019} proves the correctness
of a type checker for the Polymorphic, Cumulative Calculus of
Inductive Constructions (PCUIC), Coq's core calculus, assuming the
strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Compared to properties such as consistency, normalization, and
parametricity, syntactic type soundness is less intimidating for
dependent types since one is most unlikely to run into issues related
to the strength of the metatheory.

>>>>

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{| l |  l  | l | l | l | l| }
    \hline
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} & Main results \\
    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\

    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & TODO \\
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}

  \begin{tabular}{ll}
    \\
  Universes: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
  Inductives: & Identity types (Id), Natural numbers (Nat), W-types \\
  Conversion:& Untyped (U), Typed (T),  Extensional (E) \\
  Large Eliminations:& included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  NBE-in-Coq& \citet{nbeincoq} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
  Core Nuprl &\citet{anand2014towards} \\
  \end{tabular}

@@

Figure~\ref{fig:featurematrix} presents several mechanized proofs 
that feature logical-relations arguments for dependently-typed languages.
Each of these proofs is significantly larger than than our development; but they 
also prove more results about different object languages.
The table provides a comparison between the various features of their 
object languages, but is not exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.

@@

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} transports the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \scw{What do they prove?}
\citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further shows it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion.\scw{How did the Nuprl 
proof include a full hierarchy? Did they use impredicativity?}
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.

@@

\citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in Calculus of Constructions. We choose to
focus on a syntactic term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} proves the correctness of a type checker
for the Polymorphic, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Compared to properties such as consistency, normalization, and
parametricity, the most difficult part for syntactic type soundness is the
consistency of definitional (not propositional) equality. In their setting,
the result follows from the confluence of parallel reduction.
